---
title: "Supplementary Appendix: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013"
output: html_document
---

This is the supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013."

Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Sam Harper, PhD; Jay S Kaufman, PhD

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

```{r load_libraries, echo=FALSE, warning=F, message=F}
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
library(plotly)
library(forcats)
source("../Code/life_expectancy_functions.R")

BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")

mortality.rates <- read.csv("../Results2/mortality_rates_combined.csv")
mortality.rates.diff <- read.csv("../Results2/mortality_rates_diff_combined.csv")

mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division))) %>% rename(Race = race)
mortality.rates.diff <- mortality.rates.diff %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division))) 

#DC state abbreviation wasn't accurately defined. Adding it here.
levels(mortality.rates$stabbrs) <- c(levels(mortality.rates$stabbrs), "DC")
levels(mortality.rates.diff$stabbrs) <- c(levels(mortality.rates.diff$stabbrs), "DC")
mortality.rates$stabbrs[mortality.rates$state == "Washington DC"] <- "DC"
mortality.rates.diff$stabbrs[mortality.rates.diff$state == "Washington DC"] <- "DC"

mortality.rates <- reorder.as.map2(mortality.rates, "state", "stabbrs")
mortality.rates.diff <- reorder.as.map2(mortality.rates.diff, "state", "stabbrs")

age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
age_cod_results_infants_AOC$stabbrs[age_cod_results_infants_AOC$state == "Washington DC"] <- "DC"
age_cod_results_infants_AOC <- reorder.as.map2(age_cod_results_infants_AOC, "state", "stabbrs")

cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
```

```{r plotting_fx, echo = F}
plot.mortality.trends <- function(dataset, dataset2, COD1, sex1) {
  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
  #lim.y <- c(min(subset.data$rate.per.100k_lcl), max(subset.data$rate.per.100k_ucl))
  
  subset.data2 <- subset(dataset2, sex == sex1 & COD == COD1)
  #lim.y2 <- c(min(subset.data2$rate.difference_LCL), max(subset.data2$rate.difference_UCL))
    
  a <- ggplot(subset.data, aes(x = year, y = rate.per.100k_mean)) + 
              geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race), fill = "grey", col = NA, alpha = 0.5) +
              geom_line(aes(col = Race)) +
              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
              xlab("Year (1969-2013)") + 
              ylab("Age-standardized mortality rate (per 100,000)") + 
              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
              theme_classic(base_size = 10) +
              theme(axis.text.x = element_blank(),
                    strip.background=element_blank(),
                    axis.line=element_blank(),
                    axis.ticks=element_blank()) 
   
  b <- ggplot(subset.data2, aes(x = year, y = rate.difference_mean)) +
              geom_line() +
              geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), col = NA, alpha = 0.5) +
              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
              xlab("Year (1969-2013)") + 
              ylab("Excess mortality among blacks (per 100,000)") +
              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
              #ggtitle(paste0("Excess mortality among blacks (per 100,000) in ", sex1, " for ", COD1)) +
              theme_classic(base_size = 10) +
              theme(panel.grid.minor = element_blank(),
                    axis.text.x = element_blank(),
                    strip.background=element_blank(),
                    axis.line=element_blank(),
                    axis.ticks=element_blank()) 
  
  c <- ggplotly(a)
  d <- ggplotly(b)
  
  for(i in 1:length(c$x$data)){
    if (c$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
      c$x$data[[i]]$hoverinfo <- "none"
    }
    
    c$x$data[[i]]$line$width <- 1 #make all the lines thin
    c$x$data[[i]]$text <- gsub("rate.per.100k_mean", "Mean mortality rate", c$x$data[[i]]$text)
    c$x$data[[i]]$text <- gsub("rate.per.100k_lcl", "Lower credible limit", c$x$data[[i]]$text)
    c$x$data[[i]]$text <- gsub("rate.per.100k_ucl", "Upper credible limit", c$x$data[[i]]$text)
  }
  
  for(i in 1:length(d$x$data)){
    if (d$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
      d$x$data[[i]]$hoverinfo <- "none"
    }
    
    d$x$data[[i]]$line$width <- 1 #make all the lines thin
    d$x$data[[i]]$text <- gsub("rate.difference_mean", "Mean excess mortality", d$x$data[[i]]$text)
    d$x$data[[i]]$text <- gsub("rate.difference_LCL", "Lower credible limit", d$x$data[[i]]$text)
    d$x$data[[i]]$text <- gsub("rate.difference_UCL", "Upper credible limit", d$x$data[[i]]$text)
  }           
  
    return(list(plot1 = a, plot2 = b, plot3 = c, plot4 = d))
}
```

```{r create-the-plots, echo = F, message = F}
for(COD_i in levels(mortality.rates$COD)) {
  for(sex_i in levels(mortality.rates$sex)) {
    assign(paste0(COD_i, ".", sex_i, ".mort.plot"), plot.mortality.trends(dataset = mortality.rates, dataset2 = mortality.rates.diff,
                                                                         COD1 = COD_i, sex1 = sex_i))
  }
}
```

**Table S1: Cause of death groupings**

| Broad cause of death     | Included causes of death                                                     |
|--------------------------|------------------------------------------------------------------------------|
| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer. |
| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries. | 
| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.  |
| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death. |

**Figure S1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**

```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  labs(fill = "Census Division") +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  labs(fill = "Census Division") +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")
```  

```{r plot_LE_trends_males, echo=F, fig.height=10, fig.width=13}
LE_trend_plot_males +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure S1 Legend:** The bands illustrate the 95% credible interval around the mean trend line. States with more imprecise estimates (often due to small black populations) have wider credible intervals. The estimated life expectancy gap in California in 1972 is much higher than in surrounding years. This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error. Thus the data for this state and year should be interpreted with caution.

**Figure S2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**

```{r plot_LE_trends_females, echo=F, fig.height=10, fig.width=13}
LE_trend_plot_females +
  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)

```

```{r plot_both, eval=F, echo=F}
#Informative: male and females at the same time
ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)"))
```

**Figure S3: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**

```{r print-em, fig.width = 9, fig.height = 7, echo = F}
Cardiovascular.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Cardiovascular.Male.mort.plot[[4]]
```

**Figure S4: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**

```{r print-em-2, fig.width = 9, fig.height = 7, echo = F}
Cardiovascular.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Cardiovascular.Female.mort.plot[[4]]
```

**Figure S5: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**

```{r print-em-3, fig.width = 9, fig.height = 7, echo = F}
Cancers.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Cancers.Male.mort.plot[[4]]
```

**Figure S6: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**

```{r print-em-4, fig.width = 9, fig.height = 7, echo = F}
Cancers.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Cancers.Female.mort.plot[[4]]
```

**Figure S7: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
`Non-communicable.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
#`Non-communicable.Male.mort.plot`[[4]]
```

**Figure S8: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
`Non-communicable.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
#`Non-communicable.Male.mort.plot`[[4]]
```

**Figure S9: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
Injuries.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Injuries.Male.mort.plot[[4]]
```

**Figure S10: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
Injuries.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70)) 
#Injuries.Female.mort.plot[[4]]
```

**Figure S11: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**

```{r print-3, fig.width = 9, fig.height = 7, echo = F}
Communicable.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Communicable.Male.mort.plot[[4]]
```

**Figure S12: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**

```{r print-4, fig.width = 9, fig.height = 7, echo = F}
Communicable.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Communicable.Female.mort.plot[[4]]
```

**Figure S13: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
`All other causes.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
#`All other causes.Male.mort.plot`[[4]]
```

**Figure S14: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
`All other causes.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70)) 
#`All other causes.Female.mort.plot`[[4]]
```



**Figure S15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in males younger than one, 1969-2013**

```{r age-cod-trend-infant-males, fig.width = 12, fig.height = 10, echo = F}
ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
  geom_line(col = "#08519c") +  
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("Year (1969-2013)") +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure S16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in females younger than one, 1969-2013**

```{r age-cod-trend-infant-females, fig.width = 12, fig.height = 10, echo = F}
ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
  geom_line(col = "#08519c") +  
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("Year (1969-2013)") +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure X: Trend in the difference between each state's trend and the national trend, by causes of death and gender, 1969-2013**

```{r, fig.width = 10, fig.height = 90, echo=F}
ggplot(cod_decomp_results %>% filter(state != "Washington DC"), 
       aes(x = year, y = residual.RE)) + geom_line(aes(col = COD)) +
  facet_grid(state~sex) + geom_hline(yintercept = 0, lty = 2) +
  ylab("Difference between state and national trend (years)") +
  xlab("Year")
```

```{r, fig.width = 10, fig.height = 3, echo=F}
ggplot(cod_decomp_results %>% filter(state == "Washington DC"), 
       aes(x = year, y = residual.RE)) + geom_line(aes(col = COD)) +
  facet_grid(state~sex) + geom_hline(yintercept = 0, lty = 2) +
  ylab("Difference between state and national trend (years)") +
  xlab("Year")
```

The purpose of these plots are to summarize the information presented in Figures 3A-7B of the main paper and allow the reader to easily make comparisons between men and women and across causes of death. Each trend line shows the difference between each state's estimate and the national average for each cause of death. Thus, values close to zero are closest to the national average, values below zero imply that the state's contribution was lower than the national average and values above zero imply that the state's contribution was higher than the national average.
