---
title: 'Supplementary Appendix: Trends in the contribution of major causes of death
  to the black-white life expectancy gap by US state'
output: 
  word_document:
      reference_docx: reference.docx
---

This is supplementary appendix #1 for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state".

Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Jay S Kaufman, PhD; Sam Harper, PhD

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

For an interactive version of all our results, please visit https://corinne-riddell.shinyapps.io/black-white-life-expectancy/.

```{r load_libraries, echo=FALSE, warning=F, message=F}
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
library(plotly)
library(forcats)
source("../Code/life_expectancy_functions.R")

mortality.rates <- read.csv("../Results2/mortality_rates_combined.csv")
mortality.rates.diff <- read.csv("../Results2/mortality_rates_diff_combined.csv")

mortality.rates <- mortality.rates %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division))) %>% rename(Race = race)
mortality.rates.diff <- mortality.rates.diff %>% mutate(state.reorder = reorder(state, as.numeric(Census_Division))) 

#DC state abbreviation wasn't accurately defined. Adding it here.
levels(mortality.rates$stabbrs) <- c(levels(mortality.rates$stabbrs), "DC")
levels(mortality.rates.diff$stabbrs) <- c(levels(mortality.rates.diff$stabbrs), "DC")
mortality.rates$stabbrs[mortality.rates$state == "Washington DC"] <- "DC"
mortality.rates.diff$stabbrs[mortality.rates.diff$state == "Washington DC"] <- "DC"

mortality.rates <- reorder.as.map2(mortality.rates, "state", "stabbrs")
mortality.rates.diff <- reorder.as.map2(mortality.rates.diff, "state", "stabbrs")

age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
age_cod_results_infants_AOC$stabbrs[age_cod_results_infants_AOC$state == "Washington DC"] <- "DC"
age_cod_results_infants_AOC <- reorder.as.map2(age_cod_results_infants_AOC, "state", "stabbrs")

cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
```

```{r plotting_fx, echo = F}
plot.mortality.trends <- function(dataset, dataset2, COD1, sex1) {
  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
  #lim.y <- c(min(subset.data$rate.per.100k_lcl), max(subset.data$rate.per.100k_ucl))
  
  subset.data2 <- subset(dataset2, sex == sex1 & COD == COD1)
  #lim.y2 <- c(min(subset.data2$rate.difference_LCL), max(subset.data2$rate.difference_UCL))
    
  a <- ggplot(subset.data, aes(x = year, y = rate.per.100k_mean)) + 
              geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race), fill = "grey", col = NA, alpha = 0.5) +
              geom_line(aes(col = Race)) +
              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
              xlab("Year (1969-2013)") + 
              ylab("Age-standardized mortality rate (per 100,000)") + 
              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
              theme_classic(base_size = 10) +
              theme(axis.text.x = element_blank(),
                    strip.background=element_blank(),
                    axis.line=element_blank(),
                    axis.ticks=element_blank()) 
   
  b <- ggplot(subset.data2, aes(x = year, y = rate.difference_mean)) +
              geom_line() +
              geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), col = NA, alpha = 0.5) +
              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
              xlab("Year (1969-2013)") + 
              ylab("Excess mortality among blacks (per 100,000)") +
              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
              #ggtitle(paste0("Excess mortality among blacks (per 100,000) in ", sex1, " for ", COD1)) +
              theme_classic(base_size = 10) +
              theme(panel.grid.minor = element_blank(),
                    axis.text.x = element_blank(),
                    strip.background=element_blank(),
                    axis.line=element_blank(),
                    axis.ticks=element_blank()) 
  
  c <- ggplotly(a)
  d <- ggplotly(b)
  
  for(i in 1:length(c$x$data)){
    if (c$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
      c$x$data[[i]]$hoverinfo <- "none"
    }
    
    c$x$data[[i]]$line$width <- 1 #make all the lines thin
    c$x$data[[i]]$text <- gsub("rate.per.100k_mean", "Mean mortality rate", c$x$data[[i]]$text)
    c$x$data[[i]]$text <- gsub("rate.per.100k_lcl", "Lower credible limit", c$x$data[[i]]$text)
    c$x$data[[i]]$text <- gsub("rate.per.100k_ucl", "Upper credible limit", c$x$data[[i]]$text)
  }
  
  for(i in 1:length(d$x$data)){
    if (d$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
      d$x$data[[i]]$hoverinfo <- "none"
    }
    
    d$x$data[[i]]$line$width <- 1 #make all the lines thin
    d$x$data[[i]]$text <- gsub("rate.difference_mean", "Mean excess mortality", d$x$data[[i]]$text)
    d$x$data[[i]]$text <- gsub("rate.difference_LCL", "Lower credible limit", d$x$data[[i]]$text)
    d$x$data[[i]]$text <- gsub("rate.difference_UCL", "Upper credible limit", d$x$data[[i]]$text)
  }           
  
    return(list(plot1 = a, plot2 = b, plot3 = c, plot4 = d))
}
```

```{r create-the-plots, echo = F, message = F}
for(COD_i in levels(mortality.rates$COD)) {
  for(sex_i in levels(mortality.rates$sex)) {
    assign(paste0(COD_i, ".", sex_i, ".mort.plot"), plot.mortality.trends(dataset = mortality.rates, dataset2 = mortality.rates.diff,
                                                                         COD1 = COD_i, sex1 = sex_i))
  }
}
```

#####
**Table S1: Cause of death groupings**

| Broad cause of death     | Included causes of death                                                     |
|--------------------------|------------------------------------------------------------------------------|
| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer. |
| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries. | 
| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.  |
| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death. |

#####PAGE BREAK

**Figure S1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**

```{r print-em, fig.width = 9, fig.height = 7, echo = F}
Cardiovascular.Male.mort.plot[[1]]
#Cardiovascular.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Cardiovascular.Male.mort.plot[[4]]
```

#####PAGE BREAK

**Figure S2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**

```{r print-em-2, fig.width = 9, fig.height = 7, echo = F}
Cardiovascular.Female.mort.plot[[1]]
#Cardiovascular.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Cardiovascular.Female.mort.plot[[4]]
```

#####PAGE BREAK 
**Figure S3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**

```{r print-em-3, fig.width = 9, fig.height = 7, echo = F}
Cancers.Male.mort.plot[[1]]
#Cancers.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Cancers.Male.mort.plot[[4]]
```

#####PAGE BREAK
**Figure S4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**

```{r print-em-4, fig.width = 9, fig.height = 7, echo = F}
Cancers.Female.mort.plot[[1]]
#Cancers.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Cancers.Female.mort.plot[[4]]
```

#####PAGE BREAK
**Figure S5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
`Non-communicable.Male.mort.plot`[[1]]
#`Non-communicable.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
#`Non-communicable.Male.mort.plot`[[4]]
```

#####PAGE BREAK
**Figure S6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
`Non-communicable.Female.mort.plot`[[1]]
#`Non-communicable.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
#`Non-communicable.Male.mort.plot`[[4]]
```

#####PAGE BREAK
**Figure S7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
Injuries.Male.mort.plot[[1]]
#Injuries.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Injuries.Male.mort.plot[[4]]
```

#####PAGE BREAK
**Figure S8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
Injuries.Female.mort.plot[[1]]
#Injuries.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70)) 
#Injuries.Female.mort.plot[[4]]
```

#####PAGE BREAK
**Figure S9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**

```{r print-3, fig.width = 9, fig.height = 7, echo = F}
Communicable.Male.mort.plot[[1]]
#Communicable.Male.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Communicable.Male.mort.plot[[4]]
```

#####PAGE BREAK
**Figure S10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**

```{r print-4, fig.width = 9, fig.height = 7, echo = F}
Communicable.Female.mort.plot[[1]]
#Communicable.Female.mort.plot[[3]] %>% layout(margin = list(b = 100, l = 70))
#Communicable.Female.mort.plot[[4]]
```

#####PAGE BREAK
**Figure S11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
`All other causes.Male.mort.plot`[[1]] 
#`All other causes.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70))
#`All other causes.Male.mort.plot`[[4]]
```

#####PAGE BREAK
**Figure S12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 7, echo = F}
`All other causes.Female.mort.plot`[[1]]
#`All other causes.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100, l = 70)) 
#`All other causes.Female.mort.plot`[[4]]
```

#####PAGE BREAK
**Figure S13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in males infants (<1 year old), 1969-2013**

```{r age-cod-trend-infant-males, fig.width = 12, fig.height = 10, echo = F}
ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
  geom_line(col = "#08519c") +  
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("Year (1969-2013)") +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

#####PAGE BREAK
**Figure S14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in female infants (<1 year old), 1969-2013**

```{r age-cod-trend-infant-females, fig.width = 12, fig.height = 10, echo = F}
ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
  geom_line(col = "#08519c") +  
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("Year (1969-2013)") +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

