---
title: "Untitled"
author: "Corinne Riddell"
date: "January 31, 2017"
output: html_document
---
```{r initialize, echo=FALSE, warning=F, message=F}
library(shiny)
library(plotly)
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(purrr)
library(stringr)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")


age_cod_results <-rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"))

age_cod_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"))

age_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"))

BlackWhite_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"))

cod_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"))
```

```{r tidy_data, echo=FALSE}

ncols <- str_split_fixed(as.character(BlackWhite_results$stratum.id), "\\.", 3)
BlackWhite_results$state <- as.factor(ncols[ , 1])
BlackWhite_results$year <- as.numeric(ncols[ , 2])
BlackWhite_results$sex <- as.factor(ncols[ , 3])

age_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_decomp_results, by = "stratum.id")
cod_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], cod_decomp_results, by = "stratum.id")
age_cod_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_cod_results, by = "stratum.id")

COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
                "Non-communicable", "Injuries", "All other causes")
cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
age_cod_results$COD <- factor(age_cod_results$COD, COD.levels)
```

```{r add_Census_Region, echo=FALSE}
BlackWhite_results$Census_Region <- "South"
BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
                                                 "Nebraska", "North Dakota", "South Dakota", 
                                                 "Illinois", "Indiana", "Michigan", "Ohio", 
                                                 "Wisconsin")] <- "Midwest"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
                                                 "New Hampshire", "Rhode Island", "Vermont",
                                                 "New Jersey", "New York", "Pennsylvania")] <- "Northeast"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
                                                 "Nevada", "New Mexico", "Utah", "Wyoming",
                                                 "Alaska", "California", "Hawaii", "Oregon",
                                                 "Washington")] <- "West"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
                                                  "North Carolina", "South Carolina", 
                                                  "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "South Central"

BlackWhite_results$Census_Region <- factor(BlackWhite_results$Census_Region, levels = c("West", "Midwest", "South Central", "South Atlantic", "Northeast"))
```

```{r addvar_State_abbreviation, echo=FALSE}
BlackWhite_results <- BlackWhite_results %>% mutate(stabbrs = factor(state,  
                                                                     labels = c("AL", "AZ", "AR", "CA",
                                                                                "CO", "CT", "DE", "DC",
                                                                                "FL", "GA", "IL", "IN")))
                                      # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
                                      #            "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
                                      #            "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
                                      #            "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
                                      #            "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
                                      #            "VT", "VA", "WA", "WV", "WI", "WY")))

cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, stabbrs), 
                            cod_decomp_results, by = "stratum.id", all.y = T)
```

```{r prep_era_figure_males, echo = F}
year1 <- 1969
year2 <- 1983
year3 <- 1993
year4 <- 2013

bw_dat <- BlackWhite_results %>% 
  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap) == F) %>%
  select(state, year, LE_wbgap, sex) %>%
  arrange(year)

bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap) 
names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
bw_dat.males <- bw_dat %>% filter(sex == "Male") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
                     COD = factor(), Contribution.to.change = numeric(), 
                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())

age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
                     age_minbin = numeric(), Contribution.to.change = numeric(), 
                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())

years <- c(1969, 1983, 1993, 2013)

for (state.i in levels(BlackWhite_results$state)) {
  for (sex.i in c("Male", "Female")) {
    for (year.i in 1:3) {
      
      table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
      table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)
      
      table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
      table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)
      
      temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
                          contribution.to.gap.change(type.of.decomp = "COD", 
                                                     decomp.table1 = table1, 
                                                     decomp.table2 = table2, 
                                                     type.of.data = "results"))
      
      temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
                          contribution.to.gap.change(type.of.decomp = "Age", 
                                                     decomp.table1 = table3, 
                                                     decomp.table2 = table4,
                                                     type.of.data = "results"))
      
      cod.df <- rbind(cod.df, temp1)
      age.df <- rbind(age.df, temp2)
    }
  }
}

```

```{r more_figure_prep, echo = F}

cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state", 
                        strat.var.2 = "sex", strat.var.3 = "era", 
                        sign.var = "narrowed_gap",
                        decomp.var = "Contribution.to.change",
                        decomp.var.prop = "Contrib.to.change.prop", 
                        partition.bar.var = "COD", type.of.data = "results")

age.df <- make_pretty_decomp_plot(decomp.table = age.df, strat.var.1 = "state", 
                        strat.var.2 = "sex", strat.var.3 = "era", 
                        sign.var = "narrowed_gap",
                        decomp.var = "Contribution.to.change",
                        decomp.var.prop = "Contrib.to.change.prop", 
                        partition.bar.var = "Ages", type.of.data = "results")

cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
                                           c("state", "Census_Region", "stabbrs")], by = c("state"))
```

**Figure 1a: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r figure1, fig.align = "center", echo=F}

cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)

y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
y.num.breaks <- length(levels(factor(cod.df.males$state.gdiff.order.1)))

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
#    facet_wrap(~ Census_Region) +
    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
   # geom_segment(data = bw_dat.males,
  #               aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
  #  geom_segment(data = bw_dat.males,
  #               aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era1
```

What we see so far (based on these 12 states):

- The gap increases for some states (DC, big increase, Cali, Connecticut, and small increase in 
Colorado and Indiana). It decreases for the rest, with the largest decrease in gap seen in 
Georgia (largely due to injuries), Florida (injuries), Illinois, and Alabama/

- Injuries across the board contribute to narrowing the gap over time.

- Communicable disease narrows (except arizona), but much smaller contribution than injuries.

- All other causes narrows, except DC, again much smaller contribution than injuries.

- CVD increase gap, except in Illinois.

- Cancer increases the gap everywhere. Size of contribution to change is a little smaller than CVD.

- Non-communicable increase the gap (except Colorado) but the contribution is tiny.

**Figure 1b: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**

```{r figure1b, echo = F}
y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.males$state.gdiff.order.2)))

males.era2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era2 
```

- Be careful with the interpretation: the x-axis is much wider for this plot because of DC.

- Except for Arizona, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS). Injuries also contribute to the gap widening over this time frame. Thus, compared to previous period, these two causes of death led to a widening rather than a narrowing of the gap. 

- CVD and cancer also contributed to widening the gap over time, which also occurred in the previous period. Also similar was that non-communicable disease doesn't have much of an effect on the gap changing, while all other causes contributes to narrowing the gap.

- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period.

**Figure 1c: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**

```{r figure1c, echo = F}
y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))

males.era3 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                  ymax = state.gdiff.num.3 + 0.45, 
                  xmax = new.finish.3, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))

males.era3
```

- After 1993, the gap narrowed for all states shown, and in most cases all of the causes contributed to the narrowing of the gap. 

```{r, fig.align="center", fig.height=5, fig.width=9, echo=F, eval = F}
subplot(males.era1, males.era2, males.era3, shareY = T)
```

**Figure 2a: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r figure2a, fig.align = "center", echo=F}

cod.df.females <- merge(subset(cod.df, sex == "Female"), bw_dat.females, by = c("state"))
cod.df.females <- cod.df.females %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)

y.labels <- levels(factor(cod.df.females$state.gdiff.order.1))
y.num.breaks <- length(levels(factor(cod.df.females$state.gdiff.order.1)))

females.era1 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

females.era1
```

- The gap in life expectancy narrowed for all states shown, except for Arizona and small increase in DC.

- CVD was the biggest contributor to the gap narrowing, but all other causes and communicable diseases led to narrowing in many states. Injuries and non-communicable disease also led to narrowing, but the contribution was small. Cancer prevented the gap from narrowing further in some states.

**Figure 2b: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1983 and 1993 by US state**

```{r figure2b, fig.align = "center", echo=F}
y.labels.2 <- levels(factor(cod.df.females$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.females$state.gdiff.order.2)))

females.era2 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = second.gap, xend = second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1983-1993"), yaxis = list(title = NA, autorange = "reversed"))

females.era2
```

- Recall that during this time frame, the gap for males widened, and the largest contributors across the board were communicable disease and injuries. For females, the story is less consistent across states -- for some states, the gap widened, and for others, it narrowed. 

- Injuries did play a large role at increasing the gap in Florida, DC, and Connecticut. CVD was also generally associated with the gap increasing, while 'all other causes' was associated with narrowing the gap.

**Figure 2c: Decomposition of the change in the female black-white life expectancy gap by cause of death between 1993 and 2013 by US state**

```{r figure2c, fig.align = "center", echo=F}
y.labels.3 <- levels(factor(cod.df.females$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.females$state.gdiff.order.3)))

females.era3 <- ggplotly(
  ggplot(subset(cod.df.females, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                  ymax = state.gdiff.num.3 + 0.45, 
                  xmax = new.finish.3, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.females,
                 aes(x = fourth.gap, xend = fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
    geom_segment(data = bw_dat.females,
                 aes(x = third.gap, xend = third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))

females.era3
```

- In-line with males, the life expectancy gap decreased during the last 20 year period, and generally speaking all the causes contributed to the gap narrowing over time. The biggest contributors were CVD and all other causes.

**Figure 5: Trends in life expectancy for males between 1969 and 2013 by US state**

```{r figure5, fig.align="center", fig.height=4, fig.width=9, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_white, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(y = LE_black, col = state), lty=2) +
           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +           
           geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_wrap( ~ Census_Region) +
           ylab("Life expectancy (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs), check_overlap = T, size = 2.5) +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs, y = LE_black), check_overlap = T, size = 2.5))


```

**Figure 5: Trends in the black-white life expectancy gap for males between 1969 and 2013 by US state**

```{r figure5b, fig.align="center", fig.height=4, fig.width=9, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Male"), aes(y=LE_wbgap, x = year)) + 
           geom_vline(aes(xintercept = 1983), lty = 2, lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), lty = 2, lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_wrap( ~ Census_Region) +
           ylab("Black-white life expectancy gap (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Male"), aes(label = stabbrs), check_overlap = T, size = 2.5)) 

```

**Figure 6: Trends in life expectancy for females between 1969 and 2013 by US state**

```{r figure6a, fig.align="center", fig.height=4, fig.width=9, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_white, x = year)) + 
           geom_vline(aes(xintercept = 1983), col = "#636363", lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), col = "#636363", lwd = 0.5) +
           geom_line(aes(y = LE_black, col = state), lty=2) +
           geom_ribbon(aes(ymin = LE_black_lcl, ymax = LE_black_ucl, fill = state), alpha=0.25) +           
           geom_ribbon(aes(ymin = LE_white_lcl, ymax = LE_white_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_wrap( ~ Census_Region) +
           ylab("Life expectancy (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5) +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs, y = LE_black), check_overlap = T, size = 2.5))


```
**Figure 6: Trends in the black-white life expectancy gap for females between 1969 and 2013 by US state**

```{r figure6, fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(BlackWhite_results, sex == "Female"), aes(y=LE_wbgap, x = year)) + 
           geom_vline(aes(xintercept = 1983), lty = 2, lwd = 0.5) +
           geom_vline(aes(xintercept = 1993), lty = 2, lwd = 0.5) +
           geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) +
           facet_wrap( ~ Census_Region) +
           ylab("Black-white life expectancy gap (years)") +
           xlab("Year") +
           geom_text(data = subset(BlackWhite_results, year == 2013 & sex == "Female"), aes(label = stabbrs), check_overlap = T, size = 2.5) 
) 
```


**Figure 7a: Contribution of cancer to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7b: Contribution of cardiovascular disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7c: Contribution of communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7d: Contribution of non-communicable disease to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**

```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7d: Contribution of injuries to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 7e: Contribution of all other causes to the black-white life expectancy gap for males between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Male" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8a: Contribution of cancer to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cancers"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cancers"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8b: Contribution of cardiovascular disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Cardiovascular"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Cardiovascular"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8c: Contribution of communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8d: Contribution of non-communicable disease to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Non-communicable"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Non-communicable"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8e: Contribution of injuries to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r,  fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "Injuries"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "Injuries"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```

**Figure 8f: Contribution of all other causes to the black-white life expectancy gap for females between 1969 and 2013 by US state, in years**
```{r, fig.align="center", fig.height=5, fig.width=9, echo=F}
ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_yrs)) + 
           geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (years)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)

ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & COD == "All other causes"), aes(x = year, y = COD_cont_prop)) + 
           geom_ribbon(aes(ymin = COD_cont_prop_lcl, ymax = COD_cont_prop_ucl, fill = state), alpha=0.25) +
           geom_line(aes(col = state)) + facet_wrap(~ Census_Region) +
           ylab("Contribution to LE Gap (%)") +
           xlab("Year") +
           geom_text(data = subset(cod_decomp_results, year == 2013 & sex == "Female" & COD == "All other causes"), 
                     aes(label = stabbrs), check_overlap = T, size = 2.5)
)
```
