---
title: Historical trends in the contribution of six major causes of death to the black-white
  life expectancy gap by US state
bibliography: bibliography.bib
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

Corinne A Riddell, PhD

Kathryn T Morrison, PhD

Sam Harper, PhD

Jay S Kaufman, PhD

Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load_data, echo=F}
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
                                                                        "Communicable", "Non-communicable",  "Injuries"))
```

###Abstract

**Background**

The overall difference in life expectancy between blacks and whites in the US has decreased over time. However, states have fared differently in their progress towards eliminating the black-white life expectancy gap. Our objective is to describe the pattern of contributions of each of six major causes of deaths to the sex-specific black-white life expectancy gaps over the last half-century, and identify divergent states.

**Methods and Findings**
```{r summarize_results_for_abstract, echo = F}
cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)

ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)

can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(max = round(max(fixed.from.RE), 1))
can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))

inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
```

Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, sex, race, 19 age groups, and six major causes of death. A Bayesian autoregressive model was used to estimate smoothed mortality rates over time and Arriaga’s method was used to decompose the difference in life expectancy. Random effects meta-analysis was used to generate an overall pattern across the states, and was contrasted with the state-specific patterns to identify divergent states.

Although mortality from cardiovascular disease has decreased over time for both racial groups, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r cvd.female.2013` years). Non-communicable diseases’ contribution to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years). Both men and women exhibited a decreased contribution for injuries (men: `r inj.male.1969` to `r inj.male.2013` years, women: `r inj.female.1969` to `r inj.female.2013` years), with the contribution becoming negative for women. Several states did not follow these general trends.

**Conclusions**

The United States has shown improvement in life expectancy for both blacks and whites since 1969. For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further. In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing injury-related mortality among whites. This work emphasizes the importance of understanding how the direction of changes in mortality by race results in changes to the life expectancy gap. 


###Introduction

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethinicity. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied. 

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [@harper2007trends]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [@harper2014trends]. 

The over-arching objective of this study is to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females. This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states. This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states. 

###Methods

**Description of the mortality data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely. 

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010 [ref]. We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf].   
  
**Statistical challenges to be addressed**

The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

**Estimation of smoothed mortality rates using a Bayesian autoregressive model**

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9). 

These models were run separately for every combination of state, race, and sex. They were fit using two markov chains of length 10,000. One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation). Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles. 

**Decomposition of the life expectancy gap by cause of death**

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the model partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.

We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000 [ref].

**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**

We used random effects meta regression to describe the general pattern in these trend lines. Each state's mean contribution was regressed against a linear spline for calendar time that was estimated using four internal knots. The model included random effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year. To identify divergent states, we compared the state-specific trends (estimated separately for each state using the Bayesian model) to the general pattern estimated using meta regression.



###Results
```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")
```  

```{r plot_both, eval=F, echo=F}
#Informative: male and females at the same time
ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)"))
```

```{r plotting_resid_fx, echo = F}
plot.resid.trends <- function(dataset, COD1, sex1) {
  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
  subset.data$order <- reorder(subset.data$state, subset.data$sum.residual, first)
  
  a <- ggplot(subset.data,
         aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    #geom_line(aes(col = "Observed")) + 
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    geom_line(aes(y = fitted.FE), col = "#d7191c") +
    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
    theme_classic(base_size = 10) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
    ylab("estimated smoothed contribution vs. common trend (years)") +
    xlab("Year (1969-2013)")
  
  b <- ggplot(subset.data,
              aes(x = year, y = resid.FE)) + geom_line(aes(col = "Residual")) + 
    scale_color_manual(values = c("#d7191c")) +    
    facet_wrap( ~ order) +
    xlab(" ") + ylab("Residual") +
    geom_hline(yintercept = 0, col = "black") +
      geom_text(data = subset(subset.data, year == 2010), 
                aes(y = 1, label = round(sum.residual, 1)), size = 2.5) + 
    ggtitle(paste0("Residuals vs. year in ", COD1, " in ", sex1)) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  c <- ggplot(subset.data,
              aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
    geom_line(col = "#08519c") + 
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs. the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
    theme_classic(base_size = 10) + 
    theme(axis.text.x = element_blank(),
          strip.background=element_blank(),
          axis.line=element_blank(),
          axis.ticks=element_blank()) +
    ylab("Contribution to the life expectancy gap (years)") +
    xlab("Year (1969-2013)")
  
  return(list(plot1 = a, plot2 = b, plot3 = c))
}
```

**Men**

```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot1$x$data[[i]]$line$width <- 1
  }
}

plot1

plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot2$x$data[[i]]$line$width <- 1
  }
}

plot2
```

**Women**

```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot3$x$data[[i]]$line$width <- 1
  }
}

plot3

plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot4$x$data[[i]]$line$width <- 1
  }
}

plot4
```


```{r create-the-resid-plots, echo = F, message = F}
for(COD_i in levels(cod_decomp_results$COD)) {
  for(sex_i in levels(cod_decomp_results$sex)) {
    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.resid.trends(dataset = cod_decomp_results, 
                                                                         COD1 = COD_i, sex1 = sex_i))
  }
}
```

```{r refine_plots, eval = F, echo = F}
#need tp add the "grid lines post-hoc" because they depend on the max height of the plot


  
#turn hover off for grid lines and axis lines
refine.patterns <- function(plot) {
  for(i in 1:320){
  if(plot$x$data[[i]]$line$color == "rgba(190,190,190,1)"){
    plot$x$data[[i]]$line$width <- 0.7559055
    plot$x$data[[i]]$hoverinfo <- "none"
  }
    if(plot$x$data[[i]]$line$color == "rgba(0, 0, 0, 1)"){
    plot$x$data[[i]]$hoverinfo <- "none"
  }
  }
  return(plot)
}

#refine.patterns(Cardiovascular.Male.plot[[3]])
```


**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively. For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males. *Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women). I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South. I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancy among white women.* 

**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for males and females, in years. The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends. These lines are overlaid with the meta-estimated common trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**). 

**Cardiovascular disease**

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**). However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**). For females, black women experienced relatively larger mortality decreases over time compared to white females, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**). For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively larger than the general pattern, while Massachusettes and New York exhibit persistently lower levels of CVD contributions.

**Cancer**

For both races and both sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among males, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**). For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among males, but not females), while New York and Massachusettes underwent large decreases in the amount that cancer contributed to the life expectancy gap.

**Non-communicable disease**

Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure X**). Recently, there has been a slight decrease in mortality rates, especially among black males. These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure Y**). As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with previous causes, New York and Massachusettes exhibit decreasing contributions, and Nevada has also made substantial progress. Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap. 

**Injury**

Injury is one of the least common causes of death. Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**). For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure X**). Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases. Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among males, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among females, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

**Communicable disease**

Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communiable disease at the height of the HIV/AIDS epidemic before anti-retroviral therapy was available. This spike was greater among males, especially blacks, and within states with larger urban centers. Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among females.

**All other causes**

Mortality from all other causes generally decreased earlier in the time period, especially among blacks. White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusettes, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC. 

****

**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in males, United States, 1969-2013**

```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
LE_trend_plot_males +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in females, United States, 1969-2013**

```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
LE_trend_plot_females +
  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 3: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
Cardiovascular.Male.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 4: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-2, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cardiovascular.Female.plot[[3]]  +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-3, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cancers.Male.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) 

```

**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-4, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cancers.Female.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-5, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-6, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-7, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Injuries.Male.plot[[3]])
```

**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-8, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Injuries.Female.plot[[3]]) 
```

**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-9, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Communicable.Male.plot[[3]]) 
```

**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-10, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Communicable.Female.plot[[3]]) 
```

**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-11, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`All other causes.Male.plot`[[3]])
```

**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-12, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`All other causes.Female.plot`[[3]])
```

###Discussion 

**Summary of findings**

- In this paper, we examined state-level trends in six major CODs' contributions to the black-white LE gap.

- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result. New York and Massachusettes have virtually eliminated the black-white life expectancy gap for both men and women. While these states showed contribution decreases for all major causes of death except communicable disease, both states began with only a minimal contribution of CVD to the mortality gap.

- How do these results compare with states with large black populations: Georgia, Florida, Texas, and California? In all of these states, deaths during the first year of life are relatively higher in blacks and in most cases, deaths from CVD during mid-life is also higher among blacks. California and Florida black young men are dying more of injuries, and black men and women in Florida are dying more of communicable disease. 

- These findings can be contrasted with those of Firebaugh et al (2014). These authors distinguish between inequality caused by a difference in CVD mortality incidence vs. that caused by a difference in age-at-death and find that while blacks and whites are equally likely to die of heart disease, blacks are dying at younger ages. [qu: how can the equally likely to die of Heart disease part be true when we have the age std mort rates showing otherwise? [*See page 27 of firebaugh - highlighted section - in dropbbox papers directory*]. 

*Notes of Firebaugh 2014 paper:*
-CVD still contributes the most to the current LE gap. Firebaugh et al (2014) say that policy needs to focus on treatment once CVD is realized, not prevention ("The incidence component is negligible, indicating that blacks and whites are equally likely to succumb to heart disease. Heart disease nevertheless account for one quarter of the gap in 2015 because black victims tends to be younger than white victims. Interventions should focus on identifying heart disease and equalizing treatment. Same goes for Cancers, non-comm diseases. Contrast with HIV/AIDS, prevention is key (since diff is in "incidence", not age at death with comm disease)"
[*Does this reasoning make sense to you? I'm not sure I follow why a diff in age at death would imply focusing on treatment rather than prevention -- what if blacks get CVD at younger ages than whites? Doesn't focusing on equalizing treatment assume that blacks and whites have the same timing of onset?*]



*Literature review*

BW disparities (period-based studies)
- "higher death rates due to heart disease, cancer, homicide (men), diabetes (women), and perinatal conditions" - kochanek et al 2013, harper et al 2007; see also Orsi 2010
- harper 2012; 
- Macinko and Elo (2009): decline in disparity among working-aged people since late 1980s.

*Strengths and weaknesses*

We used data over 45 years, across two race/ethnicities, and for all age groups. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs. low *because they had decreased over time*. Furthermore, our analysis was centred on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have recieved criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ]. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles. [Methodologically, using the Bayesian autoregressive model, allowed us to estimate mortality rates in strata that were sparsely populated...*something about methods?*]

Our analysis has two main limitations. Firstly, we did not account for hispanic ethinicity, implying that both black and white populations in our study include some portion of hispanics that is known to differ over time and by state. We decided against excluding or adjusting for hispanic ethnicity in this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier time period [ref]. Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had hispanics been excluded [ref]. In a previous sensitivity analysis, we found that accounting for hispanic ethinicity did not have a substantial effect on the results [ref Harper 2012], providing some evidence that not accounting for hispanic ethnicity should not substantially affect our findings. 

Secondly, our analysis does not account for immigration, or migration between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [ref Preston 2014]. As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect life expectancy. Migration is particularly important for the District of Columbia, because of it's small geographical bounds (it is easier to be considered migration when regions are smaller - how to say this?), and substantial in-migration from New York, in- and out-migration to contiguous Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are imigrating to DC. These changes would both serve to increase the racial gap in DC.

helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
-movement of southern blacks to northern states, and migration out of rust belt.
-DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge outflux to MD and Virginia. Should almost consider these three places as one.
-lots of migraiton into MD
-NY - 24% are immigrants
-Texas has influx of immigrants to 17%
-lots of migration into virginia, as well as immigration
-Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
-Alabama gets migrants from other southern states
-Connecticut - influx in those immigrating from outside US since 1990s


Individuals who migrate are more likely to have completed a bachelor's degree than those who do not, suggesting that migration may contribute to heightened life expectancy. 

Thus, states with black or white populations that have shifted over time will be affected. This limits the interpretation of results especially for Washington DC, as it has undergone substantial demographic shifts since X. 
-foreign borns tend to have higher LE, so increase in immigration can lead to gains in LE in either blacks or whites, or both, depending on the country of origin - better because of migrant selectivity (select on traits related to good health) and better health (less likely to smoke or be obese or binge drink [ref], and return migration when they become sick)




Problems specific to DC
-migration's effect on LE over the course of the time period

-migration - has been on the decline since the 1980s (Molloy 2011, 2017). migration happens more in people with bachelor's degrees in their early twenties (and declines thereafter) but is at a bit of a higher rate throughout lifetime than those without degree (Kaplan 2017)
-roughly 1.5% of the popn moves between census regions per year, and 1.3% move within states in a census region.

Read: racial disparitiy in where people from DC move:
-https://ggwash.org/view/38326/theres-a-big-racial-disparity-in-where-people-born-in-dc-end-up-moving-or-not-moving

*Conclusions*


- DC exhibits the highest contributions for many causes, but they have the greatest gap and a widening gap - so more room for each cause to contribute. Goes to say that all causes are contributing to the gap (it isn't being driven by one cause in particular)


Life Expectancy generally
The spotlight on racial disparities in the US has intensified. Several recent papers examining US mortality by region, cause of death, or within population subgroups have received considerable attention while ignoring important black-white disparities.2, 3, 4 These disparities remain profound, and no previous paper has examined cause-specific contributions by state over an extended time period. Our analysis provides an important contribution to the literature by highlighting a high degree of heterogeneity in state-specific performance over time in eliminating these inequalities.
from above:
2. Chetty R, Stepner M, Abraham S, Lin S, Scuderi B, Turner N, Bergeron A, Cutler D. The association between income and life expectancy in the United States, 2001-2014. JAMA. 2016. 315(16):1750-1766.

3. Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al. US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014. JAMA. 2016. 316(22):2385-2401. 

4. Case A, Deaton A. Rising morbidity and mortality in midlife among white non-Hispanic Americans in the 21st century. Proc Natl Acad Sci. 2015. 112(49):15078-83. 

Cancer

Oliver. CEBP. 2008. Recent trends in BW disparities in Cancer Mortality. Note: higher mort rates in their paper likely because they restrict to >35yo.
-BW diff in lung cancer mort mimics trends in overall mort rate in males. In females, no diff in lung cancer (although minor difference in other cancer related to smoking)
-increasing disparity is seen in colorectal cancer in males between B and W
-hump also seen in prostate cancer, especially among blacks, but falling mort rates since mid 1990s.
-among females, increasing disparity in mort rates from colorectal cancer and breast cancer since 1975 (beginning of their study)
-not sure what explains the inverted-U in cancer mort in our data among females - this CEBP people doesn't seem to shed light on this.

Screening - when is it proven effective vs. when does it affect the recorded COD? Corinne to review that literature.
-breast cancer: kills mostly among those 70+ yo.

-trends in incidence may be related to overdx (prostate cancer, for ex) - for prostate cancer don't know whether the decline is due to improvements in treatment or due to screening -- could be both or just one

https://nccd.cdc.gov/uscs/toptencancers.aspx#White
-black women have higher rates of breast cancer than white women (28/100000 vs 20 for mortality)
-blacks have higher rates of prostate cancer than whites (39.1 vs 18)

Smoking and lung cancer
see: https://www.washingtonpost.com/news/to-your-health/wp/2015/11/12/smoking-among-u-s-adults-has-fallen-to-historic-lows-these-7-charts-show-who-still-lights-up-the-most/?utm_term=.f2b261e63acf -- uses CDC data for citation
-smoking rates have fallen steadily over the course of the study time period, but homogeneity in smoking still exists. used to be 42.4% in 1965 and down to 16.8% in 2014.
-by race: nearly same for blacks and whites - 17.5% in blacks vs. 18.2% in whites, but 11.2% in hispanics (which are mixed into BW designation in our study)
-by geography: highest in the midwest (20.7%), then south (17%), NE (15.3%), then west (13%), within heterogeneity within state (
Dwyer-Lindgren  L, Mokdad  AH, Srebotnjak  T, Flaxman  AD, Hansen  GM, Murray  CJ.  Cigarette smoking prevalence in US counties: 1996-2012.  Popul Health Metr. 2014;12(1):5.)

-Delancey et al (2008) identified more rapid decreases in smoking related cancers among black males since 1990 as contributing to the decrease in the overall gap.


Obesity and...
-high obesity in the south - ref 9
-women and blacks have highest rates of obesity - ref 8 

Opiods
https://www.cdc.gov/nchs/products/databriefs/db273.htm - Although rates increased 63% for non-Hispanic black persons (from 7.5 per 100,000 in 1999 to 12.2 in 2015) and 43% for Hispanic persons (from 5.4 in 1999 to 7.7 in 2015), non-Hispanic white persons experienced the greatest increase between 1999 and 2015. The age-adjusted rate of drug overdose deaths among non-Hispanic white persons increased 240% from 6.2 per 100,000 in 1999 to 21.1 in 2015.

Centers for Disease Control and Prevention. Vital signs: overdoses of prescription opioid pain relievers and other drugs among women—United States, 1999–2010. MMWR Morb Mortal Wkly Rep. 2013;62(26):537–542: opioid-related mortality increased 415% among women and 265% among men between 1999 and 2010.38

Paulozzi LJ, Xi Y. Recent changes in drug poisoning mortality in the United States by urban-rural status and by drug type. Pharmacoepidemiol Drug Saf. 2008;17(10):997–1005.: in 1999, large central metropolitan areas had the highest opioid-related mortality rates, and noncore areas had the lowest rates; by 2004, noncore areas had the highest rates and had seen the largest relative increase during that time period.50

Paulozzi LJ, Ryan GW. Opioid analgesics and rates of fatal drug poisoning in the United States. Am J Prev Med. 2006;31(6):506–511: drug poisoning mortality rates varied 8-fold between states, as did opioid consumption (methadone, 13-fold; oxycodone, 7-fold; overall, 4-fold).17 

Centers for Disease Control and Prevention. Vital signs: overdoses of prescription opioid pain relievers—United States, 1999–2008. MMWR Morb Mortal Wkly Rep. 2011;60(43):1487–1492.: Another study showed that prescription opioid sales ranged from 3.7 (Illinois) to 12.6 (Florida) kilograms per 10,000 population, and opioid-related mortality rates ranged from 5.5 (Nebraska) to 27 (New Mexico) per 100,000 population.2


-should the focus be prevention of specific condictions/diseases, treatment of those conditions/diseases, or more distal --- SES differences - edu, employment, income (Hummer Du Bois review argues we need to focus on SES distal exposures, while Firebaugh 2014 differentiates whhen to focus on treatmnet of condition vs. prevention of the condition)

-medicaid coverage varies among states. MS and NY have longer black LE than general trend and also have expanded medicare coverage (Bharmal 2012 citing the henry kaiser family foundation). CA NJ in middle for LE and coverage, AL TX lower for both -- so there is this observed ecological association



###Appendix
**S1 Table: Cause of death groupings**

| Broad cause of death     | Included causes of death                                                     |
|--------------------------|------------------------------------------------------------------------------|
| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer. |
| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries. | 
| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.  |
| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death. |


###Bibliography
