---
title: Historical trends in the contribution of six major causes of death to the black-white
  life expectancy gap by US state
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

Corinne A Riddell, PhD

Kathryn T Morrison, PhD

Sam Harper, PhD

Jay S Kaufman, PhD

Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load_data, echo=F}
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map(BlackWhite_results, "state", "stabbrs")
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/cod_decomp_results.csv")
cod_decomp_results <- reorder.as.map(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
                                                                        "Communicable", "Non-communicable",  "Injuries"))

age_cod_results_infants_AOC <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results2/age_cod_results_infants_AOC.csv")
age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
age_cod_results_infants_AOC$stabbrs[age_cod_results_infants_AOC$state == "Washington DC"] <- "DC"
age_cod_results_infants_AOC <- reorder.as.map(age_cod_results_infants_AOC, "state", "stabbrs")

```

###Abstract

**Background**

The overall difference in life expectancy between blacks and whites in the US has decreased over time. However, states have fared differently in their progress towards eliminating the black-white life expectancy gap. Our objective is to describe the pattern of contributions of each of six major causes of deaths to the sex-specific black-white life expectancy gaps over the last half-century, and identify divergent states.

**Methods and Findings**
```{r summarize_results_for_abstract, echo = F}
cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)

ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)

can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(max = round(max(fixed.from.RE), 1))
can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))

inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
```

Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, sex, race, 19 age groups, and six major causes of death. A Bayesian autoregressive model was used to estimate smoothed mortality rates over time and Arriaga’s method was used to decompose the difference in life expectancy. Random effects meta-analysis was used to generate an overall pattern across the states, and was contrasted with the state-specific patterns to identify divergent states.

Although mortality from cardiovascular disease has decreased over time for both racial groups, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r cvd.female.2013` years). Non-communicable diseases’ contribution to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years). Both men and women exhibited a decreased contribution for injuries (men: `r inj.male.1969` to `r inj.male.2013` years, women: `r inj.female.1969` to `r inj.female.2013` years), with the contribution becoming negative for women. Several states did not follow these general trends.

**Conclusions**

The United States has shown improvement in life expectancy for both blacks and whites since 1969. For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivaled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further. In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing injury-related mortality among whites. This work emphasizes the importance of understanding how the direction of changes in mortality by race results in changes to the life expectancy gap. 


###Introduction

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity. Whites lived 3.4 years longer than Blacks [ref nchs_LE]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied. 

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [@harper2007trends]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [@harper2014trends]. 

The over-arching objective of this study is to describe the pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females. This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states. This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states. 

###Methods

**Description of the mortality data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013[ref: seer stat]. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System [ref: nchs data]. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely. 

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010 [ref Census_map]. We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [ref cdc_hisp].   
  
**Statistical challenges to be addressed**

The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

**Estimation of smoothed mortality rates using a Bayesian autoregressive model**

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9). 

These models were run separately for every combination of state, race, and sex. They were fit using two Markov chains of length 10,000. One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation). Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles. 

**Decomposition of the life expectancy gap by cause of death**

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@selvin_2004; @preston]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method [@arriaga1984measuring]. This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher age-standardized cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher age-standardized rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate, say -0.5 years for injuries, this implies that whites have a higher age-standardized rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.

We also calculated age-standardized mortality rates for year-state-race-sex stratum, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000 [ref census_stdpop].

**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**

To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimate from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify divergent states, we compared the state-specific trends estimate in the previous step to the national pattern estimate by setting the random effect to zero. 



###Results
```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")
```  

```{r plot_both, eval=F, echo=F}
#Informative: male and females at the same time
ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)"))
```

```{r plotting_resid_fx, echo = F}
plot.resid.trends <- function(dataset, COD1, sex1) {
  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
  subset.data$order <- reorder(subset.data$state, subset.data$sum.residual, first)
  
  a <- ggplot(subset.data,
         aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    #geom_line(aes(col = "Observed")) + 
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    geom_line(aes(y = fitted.FE), col = "#d7191c") +
    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
    theme_classic(base_size = 10) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
    ylab("estimated smoothed contribution vs. common trend (years)") +
    xlab("Year (1969-2013)")
  
  b <- ggplot(subset.data,
              aes(x = year, y = resid.FE)) + geom_line(aes(col = "Residual")) + 
    scale_color_manual(values = c("#d7191c")) +    
    facet_wrap( ~ order) +
    xlab(" ") + ylab("Residual") +
    geom_hline(yintercept = 0, col = "black") +
      geom_text(data = subset(subset.data, year == 2010), 
                aes(y = 1, label = round(sum.residual, 1)), size = 2.5) + 
    ggtitle(paste0("Residuals vs. year in ", COD1, " in ", sex1)) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  c <- ggplot(subset.data,
              aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
    geom_line(col = "#08519c") + 
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
#    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs. the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
    theme_classic(base_size = 10) + 
    theme(axis.text.x = element_blank(),
          strip.background=element_blank(),
          axis.line=element_blank(),
          axis.ticks=element_blank()) +
    ylab("Contribution to the life expectancy gap (years)") +
    xlab("Year (1969-2013)")
  
  return(list(plot1 = a, plot2 = b, plot3 = c))
}
```

**Men**

```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot1$x$data[[i]]$line$width <- 1
  }
}

plot1

plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot2$x$data[[i]]$line$width <- 1
  }
}

plot2
```

**Women**

```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot3$x$data[[i]]$line$width <- 1
  }
}

plot3

plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot4$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot4$x$data[[i]]$line$width <- 1
  }
}

plot4
```


```{r create-the-resid-plots, echo = F, message = F}
for(COD_i in levels(cod_decomp_results$COD)) {
  for(sex_i in levels(cod_decomp_results$sex)) {
    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.resid.trends(dataset = cod_decomp_results, 
                                                                         COD1 = COD_i, sex1 = sex_i))
  }
}
```

```{r refine_plots, eval = F, echo = F}
#need tp add the "grid lines post-hoc" because they depend on the max height of the plot


  
#turn hover off for grid lines and axis lines
refine.patterns <- function(plot) {
  for(i in 1:320){
  if(plot$x$data[[i]]$line$color == "rgba(190,190,190,1)"){
    plot$x$data[[i]]$line$width <- 0.7559055
    plot$x$data[[i]]$hoverinfo <- "none"
  }
    if(plot$x$data[[i]]$line$color == "rgba(0, 0, 0, 1)"){
    plot$x$data[[i]]$hoverinfo <- "none"
  }
  }
  return(plot)
}

#refine.patterns(Cardiovascular.Male.plot[[3]])
```


**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively. For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males. 

*SH says: Don't know whether you want to note it here or not, but I find it is interesting to see that the gap is now lower in many US southern (or southeastern) states than in the Midwest, suggesting that progress has actually been somewhat greater in the South than elsewhere (particularly the upper Midwest and especially for women). I think this is interesting and somewhat counter to the narrative that health indicators are worse in the South. I also wonder how this might relate to the Murray/IHME work on US counties with "declining" life expectancy among white women.* 

**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for males and females, in years. The blue line illustrates the mean of the smoothed contribution estimated for each state-year, and illustrates the heterogeneity across states in their contribution trends. These lines are overlaid with the meta-estimated common trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**). 

**Cardiovascular disease**

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**). However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**). For females, black women experienced relatively larger mortality decreases over time compared to white females, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**). For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively larger than the general pattern, while Massachusetts and New York exhibit persistently lower levels of CVD contributions.

**Cancer**

For both races and both sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among males, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**). For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among males, but not females), while New York and Massachusetts underwent large decreases in the amount that cancer contributed to the life expectancy gap.

**Non-communicable disease**

Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure 5**). Recently, there has been a slight decrease in mortality rates, especially among black males. These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure 6**). As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with previous causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress. Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap. 

**Injury**

Injury is one of the least common causes of death. Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure 7**). For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure 8**). Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases. Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among males, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among females, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

**Communicable disease**

Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**). This spike was greater among males, especially blacks, and within states with larger urban centers. Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among females.

**All other causes**

```{r, echo = F}
male.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Male", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

female.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Female", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

male.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

female.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
```


Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**eFigures 11 and 12**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC. 

*CR note: I just added the following paragraph. Open to how/whether we want to talk about this. Thought we need to say something about how "all other causes partiallly reflects infant mortality. Also these additional figures (Fig 15 and 16) would be moved to the appendix.*

To investigate how much of the gap in all other causes occurred during the first year of life, we conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred *during the first year of life* to the life expectancy gap (**Figures 15 and 16**). We investigated this because the majority of infant deaths are classified as being due to all other causes, conflating the all other causes category with inquities in infant mortality. We found that 43% of the contribution of all other causes to the gap was rooted in differences during the first year of life. In 2013, 53% of these deaths occurred during the first year of life. Thus, the trend in the all other causes contribution partially reflects a decreasing disparity in infant mortality.

****

**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in males, United States, 1969-2013**

```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
LE_trend_plot_males +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 15), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in females, United States, 1969-2013**

```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
LE_trend_plot_females +
  geom_hline(aes(yintercept = -5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 10), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 3: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
Cardiovascular.Male.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 4: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-2, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cardiovascular.Female.plot[[3]]  +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-3, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cancers.Male.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) 

```

**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-4, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
Cancers.Female.plot[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-5, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-6, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

```

**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-7, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Injuries.Male.plot[[3]])

`Injuries.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
           geom_hline(aes(yintercept = 6), lwd = 0.5, col = "grey", alpha = 0.5) 
  
```

**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-8, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Injuries.Female.plot[[3]]) 

`Injuries.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-9, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Communicable.Male.plot[[3]]) 

`Communicable.Male.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) 
```

**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-10, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Communicable.Female.plot[[3]])

`Communicable.Female.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-11, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(`All other causes.Male.plot`[[3]])

`All other causes.Male.plot`[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) 
```

**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-12, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(`All other causes.Female.plot`[[3]])

`All other causes.Female.plot`[[3]] +
  geom_hline(aes(yintercept = -2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure 15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in males younger than one**

```{r age-cod-trend-infant-males, fig.width = 9, fig.height = 9, echo = F}
ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Male"), aes(y = age_COD_cont_yrs_mean, x = year)) +
  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
  geom_line(col = "#08519c") +  
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("Year (1969-2013)") +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

**Figure 16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the overall pattern in females younger than one**

```{r age-cod-trend-infant-females, fig.width = 9, fig.height = 9, echo = F}
ggplot(dat = subset(age_cod_results_infants_AOC, sex == "Female"), aes(y = age_COD_cont_yrs_mean, x = year)) +
  geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = age_COD_cont_yrs_lcl, ymax = age_COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
  geom_line(col = "#08519c") +  
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (years)") +
  xlab("Year (1969-2013)") +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```


###Discussion 

**Summary of findings**

In this paper, we examined state-level trends in six major CODs' contributions to the black-white life expectancy gap. While we found that many states mimicked the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result. 

We found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries continue to play a role in contributing to the current gap for males in several states, especially in the Midwest and DC, while contributing to a decreased gap among females in the South. Illinois, Michigan and Wisconsin also show relatively high contributions from cancer. On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years. Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing minimal difference and point to the importance of viewing the most recent estimates in light of their earlier values.

**Literature review** 

This research extends previous studies that have estimated either the contribution of causes of death to the life expectancy gap on the national level, or the excess risk among blacks in cause of death specific mortality rates [@harper2007trends; @harper2012trends; orsi2010black] [ref kochanek_2013]. 

*Strengths and weaknesses*

We used data over 45 years, across two race/ethnicities, and for all age groups. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs. low *because they had decreased over time*. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ]. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.

Our analysis has two main limitations. Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state. We decided against excluding or adjusting for Hispanic ethnicity in this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier time period [ref]. Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [ref]. In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [ref Harper 2012], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings. 

Secondly, our analysis does not account for immigration, or migration between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [ref Preston 2014]. As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect the estimated life expectancy of a state. Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Such changes would both serve to increase the racial gap in DC.

Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.  

**Conclusions** 

This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among state in risk factors leading to each cause of death. Differences in smoking policy, medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state. 


###Appendix
**S1 Table: Cause of death groupings**

| Broad cause of death     | Included causes of death                                                     |
|--------------------------|------------------------------------------------------------------------------|
| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer. |
| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries. | 
| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.  |
| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death. |


###Bibliography


to add to bibliography list:
nchs_LE ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf

seer stat: National Cancer Institute. Surveillance, Epidemiology, and End Results
Program: SEER*Stat software: latest release: version 8.1.5 [Internet]. Bethesda (MD): NCI; 2013 Mar 31 [cited 2014 Jun 20]. Available from: http://www.seer.cancer.gov/seerstat

nchs data: National Cancer Institute. Surveillance, Epidemiology, and End Results Program: US mortality data [Internet]. Bethesda (MD): NCI; [cited 2014 Jun 20]. Available from: http://www.seer.cancer.gov/mortality/

cdc_hisp: [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf]

census_map: https://www.census.gov/2010census/popmap/. retrieved apr 25, 2017

census_std_pop: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html. retrieved apr 25, 2017

kochanek_2013: https://www.cdc.gov/nchs/data/databriefs/db125.pdf