---
title: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

Corinne A Riddell, PhD

Kathryn T Morrison, PhD

Jay S Kaufman, PhD

Sam Harper, PhD

All authors: Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 27, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>. Phone: +1-438-863-0402

```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

source("../Code/life_expectancy_functions.R")
```

```{r load_data, echo=F}
cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
                                                                        "Communicable", "Non-communicable",  "Injuries"))

age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
```

###Abstract

Life expectancy has increased in the United States over many decades. The difference in life expectancy between black and white Americans has also decreased, but some states have made much more progress towards racial equality than others. This paper describes the pattern of contributions of six major causes of death to the black-white life expectancy gap within US states and the District of Columbia between 1969 and 2013, and identifies states diverging from the overall pattern. Across multiple causes, the District of Columbia, Illinois, Wisconsin, and Michigan had the highest contributions to black-white inequality, while New York, Massachusetts, and Rhode Island had the lowest contributions and have either achieved or are the closest to achieving black-white equality in life expectancy. 

----

###Introduction

Life expectancy is a barometer for overall population health. In 2014, life expectancy at birth in the United States (US) was 78.8 years, but hidden within this summary are important subgroup differences. Whites lived 3.4 years longer than blacks (Cunningham et al., 2017). While both black and white Americans have experienced increases in life expectancy since the beginning of the twentieth century, the difference between their life expectancies has varied over time (Fuchs, 2016). 

During the 1980s the life expectancy gap widened, but then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths from homicide, HIV/AIDS, unintentional injuries, and, among women, from differential improvements in heart disease (Harper et al., 2007). The overall black-white gap continued to narrow during the 2000s, but US states differed markedly in the magnitude of improvement. For example, Harper et al. estimated that the black-white life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years (Harper et al., 2014). While the previous publication estimated changes to the black-white life expectancy gap for each US state, it did not shed light on why some states progressed more than others.

Race-specific mortality trajectories within a state are likely affected by both national and state-specific factors, some of which could be affected by state-level policymaking. Taking cardiovascular disease (CVD) as an example, several state policies have direct or indirect effects on CVD risk factors. For example, excise taxes on tobacco were introduced between 1921 and 1969 across all states, and the magnitude of increases have varied across states, leading to large present-day differences (e.g., \$0.17 in Missouri vs. \$4.35 in New York) (Centers for Disease Control and Preventaion, 2018). Policies targeting obesity through promotion of physical activity and nutrition also vary by state, (Trust for America's Health and the Robert Wood Johnson Foundation, 2018) as do restrictions on vehicle emissions, with twelve states following the stricter standards imposed in California (California Air Resources Board, 2018). Thus, policies and other state-level characteristics may be implicated in deaths from specific causes. However, a first step in understanding the potential impact of state policies is to examine the extent to which different causes of death have contributed to state-level changes in the black-white life expectancy gap.

The overarching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women. This facilitates the identification of causes of death whose mortality trends for blacks and whites have changed in ways that contributed to narrowing or widening the black-white life expectancy gap, and the identification of states for which cause-specific contributions are particularly high or low compared to the overall average. The study may therefore facilitate the identification of policies or social conditions states with narrow black-white gaps, and suggest broad targets for interventions to further reduce the black-white gap in less states that have seen less progress in reducing racial inequality.

###Methods

**Description of the mortality data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in 39 US states and the District of Columbia between 1969 and 2013.^5^ We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010.^6^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^7^ The population data were derived by the NCHS using intercensal bridged-race population estimates. We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**). We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^8^ 

**Statistical challenges**

The data structure produced two statistical challenges that were addressed by our modeling strategy. Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ which we needed to impute. Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata. In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.

**Estimation of smoothed mortality rates using a Bayesian time-series model**

We used a Bayesian time-series model to smooth the death counts across years and estimate the underlying mortality rate. Specifically, we used a Bayesian generalized linear model with a Poisson likelihood to model the number of deaths in each state-year-sex-race-age-cause stratum as a function of the at-risk population, for strata with zero or more than 9 deaths. The mortality rates over time within each stratum were assigned an autoregressive prior, where the log mortality rate was assumed to be Gaussian centered at the previous year's mortality rate. A non-informative gamma prior was used to estimate the inverse variance. For censored strata with between one and nine deaths, the deaths were modeled using a truncated Poisson likelihood, and otherwise the model remained the same. The truncated Poisson was bound between one and the larger of nine or the stratum's population size, which could be less than nine. 

Models were fit using Markov chain Monte Carlo (MCMC). These models were run separately for every combination of state, race, gender, and cause of death. They were fit using two chains of length 10,000, with a 2,000 iteration burn-in period to ensure convergence. One thousand iterations remained after burn-in and sampling every eighth iteration to reduce autocorrelation among the MCMC samples. Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles. This modeling was done in R (version 3.3.2) using the *rjags* package.

**Decomposition of the life expectancy gap by cause of death**

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total number of deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^12^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate this implies that whites have a higher rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.

We also calculated state-level trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^13^ The calculations described in this section were performed using functions written by the first author.

**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**

To describe the national trend in each cause's contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a cubic spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero. These estimates relied on the R *lme4* package for random effects modeling, and the *splines* package to define cubic splines.

A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: link [DOI: removed for double blind review]

###Results

```{r plotting_contributions_fx, echo = F}
plot.contribution.trends <- function(dataset, COD1, sex1) {
  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
  subset.data$order <- reorder(subset.data$state, subset.data$sum.residual, first)
  
  a <- ggplot(subset.data,
         aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    #geom_line(aes(col = "Observed")) + 
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    geom_line(aes(y = fitted.FE), col = "#d7191c") +
    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
    theme_classic(base_size = 14) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
    ylab("estimated smoothed contribution vs. common trend (years)") +
    xlab("Year (1969-2013)")
  
  b <- ggplot(subset.data,
              aes(x = year, y = resid.FE)) + geom_line(aes(col = "Residual")) + 
    scale_color_manual(values = c("#d7191c")) +    
    facet_wrap( ~ order) +
    xlab(" ") + ylab("Residual") +
    geom_hline(yintercept = 0, col = "black") +
      geom_text(data = subset(subset.data, year == 2010), 
                aes(y = 1, label = round(sum.residual, 1)), size = 2.5) + 
    ggtitle(paste0("Residuals vs. year in ", COD1, " in ", sex1)) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  c <- ggplot(subset.data,
              aes(x = year, y = COD_cont_yrs_mean)) + 
    geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + #geom_hline(yintercept = 0, col = "black") +
    geom_line(aes(y = fitted.RE - RE.estimates), col = "#d7191c") +
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9", alpha = 0.65) +
    geom_line(col = "#08519c") + 
    #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display    
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
    #geom_line(aes(y = fitted.RE), col = "#d7191c", lty = 3) +
#    ggtitle(paste0("Estimated smoothed contribution of ", COD1, " disease to the black-white life expectancy gap (blue 95% credible interval) vs. the meta-estimated general pattern (red line) in ", sex1, " 1969-2013")) +
    theme_classic(base_size = 14) + 
    theme(axis.text.x = element_blank(),
          strip.background=element_blank(),
          axis.line=element_blank(),
          axis.ticks=element_blank()) +
    ylab("Contribution to the life expectancy gap (years)") +
    xlab("Year (1969-2013)")
  
  return(list(plot1 = a, plot2 = b, plot3 = c))
}
```

```{r create-the-resid-plots, echo = F, message = F}
for(COD_i in levels(cod_decomp_results$COD)) {
  for(sex_i in levels(cod_decomp_results$sex)) {
    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.contribution.trends(dataset = cod_decomp_results, 
                                                                         COD1 = COD_i, sex1 = sex_i))
  }
}
```

```{r refine_plots, eval = F, echo = F}
#need tp add the "grid lines post-hoc" because they depend on the max height of the plot


  
#turn hover off for grid lines and axis lines
refine.patterns <- function(plot) {
  for(i in 1:320){
  if(plot$x$data[[i]]$line$color == "rgba(190,190,190,1)"){
    plot$x$data[[i]]$line$width <- 0.7559055
    plot$x$data[[i]]$hoverinfo <- "none"
  }
    if(plot$x$data[[i]]$line$color == "rgba(0, 0, 0, 1)"){
    plot$x$data[[i]]$hoverinfo <- "none"
  }
  }
  return(plot)
}

#refine.patterns(Cardiovascular.Male.plot[[3]])
```

**Figure S1 and S2** display trends in the life expectancy gap for men and women, respectively. For men and women, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men. 

**Figures 1 - 6** display trends in the contribution of each broad cause of death to the life expectancy gap for men (panel A) and women (panel B), in years, for each state. The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines. These lines are overlaid with the meta-estimated national trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S3-S14**). 

**Cardiovascular disease**
```{r summarize_results, echo = F}
cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>%
  ungroup() %>% select(fixed.from.RE) %>% round(1)
cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)

ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>%  select(fixed.from.RE) %>% round(1)
ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)

can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))

inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
```

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years (**Figure 1A**). Black women experienced relatively larger CVD mortality decreases over time compared to white women, resulting in a decrease of CVD to the contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 1B**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.


**Figure 1A: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot, fig.width = 12, fig.height = 10, echo = F}
cvd.men <- Cardiovascular.Male.plot[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
  ggtitle("A) Contribution of CVD to the gap in men")

```

**Figure 1 Legend:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time. The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states. The early spike in California's trend line illustrates the unexpectedly high estimate of the CVD contribution to the black-white life expectancy gap among California males in 1972. This anomolous estimate reflects the unexpectedly low CVD-related mortality rate among white males in California in 1972, and likely denotes a coding error. Thus the data for this state and year should be interpreted with caution.

**Figure 1B: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-2, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
cvd.women <- Cardiovascular.Female.plot[[3]]  +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
  ggtitle("B) Contribution of CVD to the gap in women")

```

**Cancer**

For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**). Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**), with a peak contribution of `r can.male.max` in the early 1990s. For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a generally stable national pattern in the contribution of cancer to the life expectancy over time of approximately `r can.female.average` years (**Figure 2B**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.

**Figure 2A: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-3, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
can.men <- Cancers.Male.plot[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) + 
  ggtitle("A) Contribution of cancer to the gap in men")

```

**Figure 2B: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-4, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
can.women <- Cancers.Female.plot[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) + 
  ggtitle("B) Contribution of cancer to the gap in women")

```

**Non-communicable disease**

Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S7**). Recently, there was a slight decrease in mortality rates, especially among black men. These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**). As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men (approximately `r ncd.male.average` years), and exhibited a gradual decrease over time among women from `r ncd.female.1969` to `r ncd.female.2013` years (**Figures 3A and 3B**). As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress. Furthermore, while DC's mortality rates decreased, they presently have the highest contribution of non-communicable disease to the gap. 

**Figure 3A: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-5, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ncd.men <- `Non-communicable.Male.plot`[[3]] +
  geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  scale_y_continuous(breaks = seq(-0.5, 1.5, by = 0.5), labels = seq(-0.5, 1.5, by = 0.5)) +
  ggtitle("A) Contribution of non-communicable disease to the gap in men")

```

**Figure 3B: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-6, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ncd.women <- `Non-communicable.Female.plot`[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  scale_y_continuous(breaks = seq(-1, 1.5, by = 0.5), labels = seq(-1, 1.5, by = 0.5)) + 
  ggtitle("B) Contribution of non-communicable disease to the gap in women")

```

**Injury**

Trends in mortalilty from injuries differ for blacks and whites. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**). Thus, injury mortality rates among blacks are equivalent to, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore decreased substantially over time from `r inj.male.1969` to `r inj.male.2013` years among men. Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.


**Figure 4A: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-7, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Injuries.Male.plot[[3]])

inj.men <-`Injuries.Male.plot`[[3]] +
  geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 6), lwd = 0.5, col = "grey", alpha = 0.5) + 
  ggtitle("A) Contribution of injury to the gap in men")
  
```

**Figure 4B: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-8, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Injuries.Female.plot[[3]]) 

inj.women <- `Injuries.Female.plot`[[3]] +
  geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  ggtitle("B) Contribution of injury to the gap in women")
```


**Communicable disease**

```{r communicable_dx_stats, echo=F}
com.male.1969 <- cod_decomp_results %>% filter(year == 1969, sex == "Male" & COD == "Communicable" & state == "Alabama") %>% 
  ungroup() %>% select(fixed.from.RE) %>% round(1)

com.female.1969 <- cod_decomp_results %>% 
  filter(year == 1969, sex == "Female" & COD == "Communicable" & state == "Alabama") %>% 
  ungroup() %>% select(fixed.from.RE) %>% round(1)

com.male.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  ungroup() %>%
  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))

com.female.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  ungroup() %>%
  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))

```

Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**). This spike was greater among men, especially blacks, and among states with larger urban centers. The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the early 1990s on `r com.male.mid.max` years in men and `r com.female.mid.max` years in women, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.

**Figure 5A: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-9, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Communicable.Male.plot[[3]]) 

com.men <- `Communicable.Male.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  ggtitle("A) Contribution of communicable disease to the gap in men")
```

**Figure 5B: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-10, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Communicable.Female.plot[[3]])

com.women <- `Communicable.Female.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  ggtitle("B) Contribution of communicable disease to the gap in women")
```
**All other causes**

```{r, echo = F}

aoc.male.endpoints <- cod_decomp_results %>% 
  filter(year %in% c(1969, 2013), sex == "Male" & COD == "All other causes" & state == "Alabama") %>% 
  ungroup() %>% select(fixed.from.RE) %>% round(1)

aoc.female.endpoints <- cod_decomp_results %>% 
  filter(year %in% c(1969, 2013), sex == "Female" & COD == "All other causes" & state == "Alabama") %>% 
  ungroup() %>% select(fixed.from.RE) %>% round(1)

aoc.male.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)

aoc.female.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)

proportions.males <- aoc.male.endpoints.infant/aoc.male.endpoints
proportions.females <- aoc.female.endpoints.infant/aoc.female.endpoints
```

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S13 and S14**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time (from `r sprintf("%.1f", aoc.male.endpoints[1, 1])` to `r aoc.male.endpoints[2, 1]` years among males and `r aoc.female.endpoints[1, 1]` to `r aoc.female.endpoints[2, 1]` years among females), and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**). The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC. 

Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally. We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**). We found that `r proportions.males[1,1]*100`% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969 in males and `r round(proportions.females[1,1]*100, 0)`% in females. In 2013, this proportion reached `r proportions.males[2,1]*100`% in males and `r proportions.females[2,1]*100`% in females. Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.

**Figure 6A: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

```{r print-plot-11, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(`All other causes.Male.plot`[[3]])

aoc.men <- `All other causes.Male.plot`[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) +
  ggtitle("A) Contribution of all other causes to the gap in men")
```

**Figure 6B: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

```{r print-plot-12, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(`All other causes.Female.plot`[[3]])

aoc.women <- `All other causes.Female.plot`[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) +
  ggtitle("B) Contribution of all other causes to the gap in women")
```

```{r save-figs}
ggsave(filename = "../Plots/figure2a.jpeg", plot = cvd.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure2b.jpeg", plot = cvd.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure3a.jpeg", plot = can.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure3b.jpeg", plot = can.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure4a.jpeg", plot = ncd.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure4b.jpeg", plot = ncd.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure5a.jpeg", plot = inj.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure5b.jpeg", plot = inj.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure6a.jpeg", plot = com.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure6b.jpeg", plot = com.women, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure7a.jpeg", plot = aoc.men, device = "jpeg", width = 12, height = 10, units = "in")
ggsave(filename = "../Plots/figure7b.jpeg", plot = aoc.women, device = "jpeg", width = 12, height = 10, units = "in")
```

###Discussion 

**Summary of findings**

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period. While we found that many states conformed to a general pattern, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, because greater racial equality can be achieved by either falling black mortality rates or rising white rates, the reduced life expectancy gap cannot be celebrated without a careful examination of the two individual trends.^16^

Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida. In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV. Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts). Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years. Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.

**Strengths and weaknesses**

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum. In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities with divergent social environments that are reflected in varying patterns of mortality by race.

Our analysis has three main limitations. Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years. Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have lower mortality than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of its small geographical bounds. Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^

**Implications for Policy & Future Work**

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time. States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others. Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^

###Bibliography

1. Cunningham TJ, Croft JB, Liu Y, Lu H, Eke PI, Giles WH (2017) Vital Signs: Racial Disparities in Age-Specific Mortality Among Blacks or African Americans - United States, 1999-2015. *MMWR Morb Mortal Wkly Rep* 66(17):444-456.

2. Fuchs, VR (2016) Black Gains in Life Expectancy. *JAMA*. 316(18): 1869–70. 

3. Harper S, Lynch J, Burris S, Smith GD (2007) Trends in the black-white life expectancy gap in the United States, 1983-2003. *JAMA* 297(11):1224-32.

4. Harper S, MacLehose RF, Kaufman JS (2014) Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009. *Health Aff* 33(8): 1375–82.

5. National Cancer Institute. Surveillance, Epidemiology, and End Results
Program: SEER*Stat software: release: version 8.3.2 [Internet]. Bethesda (MD): NCI; 2017 Apr 14 [cited 2016 August 20]. Available from: http://www.seer.cancer.gov/seerstat.

6. U.S. Census Bureau: United States Census 2010 Interactive Population Map [Internet]. [cited 2017 May 16]. Available from: https://www.census.gov/2010census/popmap/.

7. Centers for Disease Control and Prevention. National Center for Health Statistics: National Vital Statistics System [Internet]. [cited 2017 May 16]. Available from: https://www.cdc.gov/nchs/nvss/index.htm.

8. Arias E, Schauman WS, Eschbach K, Sorlie PD, Backlund E (2008) The validity of race and Hispanic origin reporting on death certificates in the United States. National Center for Health Statistics. Vital Health Stat 2(148). Available from: https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf

9. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: US mortality data [Internet]. Bethesda (MD): NCI; [cited 2017 May 16]. Available from: http://www.seer.cancer.gov/mortality/

10. Preston SH, Heuveline P, and Guillot M (2001) Demography: Measuring and Modeling Population Processes. (Blackwell Publishing, Malden)

11. Selvin, S (2004) Statistical Analysis of Epidemiologic Data. 3rd ed. (Oxford University Press, Oxford)

12. Arriaga EE (1985) Measuring and explaining the change in life expectancies. *Demography* 21(1):83-96.

13. National Cancer Institute. Surveillance, Epidemiology, and End Results Program: Standard Populations - 19 Age Groups [Internet]. [cited 2017 April 25]. Available from: https://seer.cancer.gov/stdpopulations/stdpop.19ages.html. 

14. Lemann, N (1991) The Promised Land: The Great Black Migration and How It Changed America. New York: Alfred A. Knopf, ISBN 0-394-56004-3.

15. Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al (2016) US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014. *JAMA* 316(22):2385–2401.

16. Case A, Deaton A (2017) Mortality and Morbidity in the 21st Century. In: Eberly J and Stock J, Editors. Brookings Papers on Economic Activity, March 23-24 2017.

17. Harper S, Rushani D, and Kaufman JS (2012) Trends in the Black-White Life Expectancy Gap, 2003-2008. *JAMA* 307(21):2257–9.

18. Rudd RA, Seth P, David F, Scholl L (2016) Increases in Drug and Opioid-Involved Overdose Deaths — United States, 2010–2015. *MMWR Morb Mortal Wkly Rep* 65:1445–1452.

19. Case A, Deaton A (2015) Rising Morbidity and Mortality in Midlife among White Non-Hispanic Americans in the 21st Century. *Proc Natl Acad Sci * 112(49):15078-83. 

20. Minton J, Green M, McCartney G, Shaw R, Vanderbloemen L, Pickett K (2016) Two cheers for a small giant? Why we need better ways of seeing data: A commentary on:‘Rising morbidity and mortality in midlife among White non-Hispanic Americans in the 21st century’. *Int J Epidemiol* 2016 Sep 17:dyw095.

21. Arora A, Spatz E, Herrin J, Riley C, Roy B, Kell K, Coberley C, Rula E, Krumholz HM (2016) Population Well-Being Measures Help Explain Geographic Disparities In Life Expectancy At The County Level. *Health Aff* 35(11):2075-82.

22. Ruiz JM, Steffen P, Smith TB (2013) Hispanic mortality paradox: a systematic review and meta-analysis of the longitudinal literature. *Am J Public Health* 103(3):e52-60.

23. Preston SH, Elo IT (2014) Anatomy of a municipal triumph: New York City's upsurge in life expectancy. *Population and development review* 40(1):1-29.

24. U.S. Census Bureau. Census Reporter Profile Page: American Community Survey 1-year estimates [Internet]. [Cited 16 May 2017]. Available from: https://censusreporter.org/profiles/16000US1150000-washington-dc/.

25. Kuehn BM (2014) CDC: major disparities in opioid prescribing among states: some states crack down on excess prescribing. *JAMA* 312(7):684-6.

26. Jones CM, Mack KA, Paulozzi LJ (2010) Pharmaceutical overdose deaths, United States, 2010. *JAMA* 309(7):657-9.

27. Cotti C, Nesson E, Tefft N (2016) The Effects of Tobacco Control Policies on Tobacco Products, Tar, and Nicotine Purchases among Adults: Evidence from Household Panel Data. *American Economic Journal: Economic Policy* 8(4):103-23.

28. Grogan CM, Park SE (2017) The Racial Divide in State Medicaid Expansions. *J Health Polit Policy Law* 42(3):539-72.

29. Humphreys DK, Gasparrini A, Wiebe DJ (2017) Evaluating the Impact of Florida’s “Stand Your Ground” Self-defense Law on Homicide and Suicide by Firearm: An Interrupted Time Series Study. *JAMA Intern Med* 177(1):44-50.

30. Taber DR, Stevens J, Evenson KR, Ward DS, Poole C, Maciejewski ML, Murray DM, Brownson RC (2011) State policies targeting junk food in schools: racial/ethnic differences in the effect of policy change on soda consumption. *Am J Public Health* 101(9):1769-75.

31. Adimora AA, Ramirez C, Schoenbach VJ, Cohen MS (2014) Policies and politics that promote HIV infection in the Southern United States. *AIDS*
28(10):1393.

32. Strumpf EC, Harper S, Kaufman JS (2017) Fixed Effects and Difference in Differences. In: Methods in Social Epidemiology. 2nd Ed, eds Oakes M, Kaufman JS(Jossey-Bass, San Francisco).  

###Footnotes

**Contributors**

CAR: Extracted the data, performed data analysis, drafted the manuscript.

KTM: Performed data analysis, reviewed the paper for intellectual content.

SH: Initial conception for the study, interpreted the results, reviewed the paper for intellectual content.

JK: Initial conception for the study, interpreted the results, drafted the manuscript, reviewed the paper for intellectual content.

We extend our thanks to Dr. Richard Maclehose, who developed the initial version of the code to perform the Bayesian analysis as part of the research article “Trends in the Black-White Life Expectancy Gap Among Us States, 1990–2009”, by Harper S, MacLehose RF, and Kaufman JS.

**Funding**

This research received no specific grant from any funding agency in the public, commercial or not-for-profit sectors. 

Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr. Kaufman's Canada Research Chair in Health Disparities. Kathryn Morrison is a post-doctoral researcher at McGill University. Sam Harper was supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec. Jay Kaufman was supported by a Canada Research Chair in Health Disparities. 

**Competing Interests**

All authors have completed the ICMJE uniform disclosure form at www.icmje.org/coi_disclosure.pdf and declare: no support from any organisation for the submitted work; no financial relationships with any organisations that might have an interest in the submitted work in the previous three years; no other relationships or activities that could appear to have influenced the submitted work.”
 
**Data Sharing Statement**

A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: https://github.com/corinne-riddell/BlackWhiteMortalityGap/ (DOI: 10.5281/zenodo.803472).
